Describing *Symbiodinium* metacommunities using high-throughput sequencing

### 2017-02-08 manuscript/manuscript.Rmd
Pete revisions round 3

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-tolerance of their coral hosts, making the description of *Symbiodinium* communities in corals (and metacommunities on reefs) a routine and essential goal in coral reef research. High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented scale, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity. Using samples of *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St. John, US Virgin Islands, we demonstrate that generating operational taxonomic units by clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific. We utilized this ‘within-sample clustering’ approach to create metacommunity networks of host-symbiont associations in St. John, which revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations with *Symbiodinium*, although no differences in symbiont communities were observed between north and south shore environments. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

@@

  *Symbiodinium* identity is primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; @Santos:2003kt; @Pochon:2012we]. Together, these markers have been used to identify nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which have been further divided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, ecological surveys of *Symbiodinium* diversity must still rely on commonly-used marker genes, such as ITS2. Consequently, high-throughput sequencing of ITS2 increasingly is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs]. However, such analyses may create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within the genus [@Finney:2010p8324; @Thornhill:2013dx]. For example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially other species [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly-repeating ribosomal DNA array creates multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (e.g. [@Dover:1986ju]). In fact, concerted evolution may mask species divergence within *Symbiodinium* by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker for *Symbiodinium*. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level. Moreover, due to the large quantity of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2, it remains an essential and powerful marker for the *Symbiodinium* system. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys of *Symbiodinium*.

  Here, we describe a *Symbiodinium* metacommunity associated with scleractinians and a *Millepora* hydrocoral in St. John, U.S. Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that may advance understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' (i.e. clustering sequences at 97% similarity within each sample independently) to generate biologically relevant operational taxonomic units (OTUs) from ITS2 metabarcoding data. Specifically, this approach addresses the fact that dominant ITS2 sequences from different *Symbiodinium* species may be more similar to one another than intragenomic variants within one *Symbiodinium* [@Thornhill:2013dx; @Arif:2014gk; @Parkinson:2015gt]. Therefore, clustering all sequences together underestimates diversity by collapsing different species into the same OTU, while treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to intragenomic variation. Within-sample clustering may better approximate true diversity by exploiting key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts. These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to be intragenomic variants, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa. Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using within-sample clustering, we analyzed *Symbiodinium* communities in ten host species across the north and south shores of St. John. We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying the variability in symbiont communities (i.e., beta diversity, or 'symbiosis flexibility') among individuals within a host species. The large quantity of data, and the kinds of analytical approaches facilitated by ITS2 metabarcoding, have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts. More generally, the present study demonstrates how ecological knowledge can inform bioinformatic analyses using marker genes for which sequence diversity does not map directly to species.

@@

  Eighty-four tissue samples were collected from scleractinian and milleporine corals sampled at seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and all colonies to be sampled were selected haphazardly in upward-facing locations exposed to ambient irradiance. Tissue biopsies (~5-mm diameter and ~5-mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.

  Broad-scale differences in the hydrographic regimes at points representative of the north and south shores of St. John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and integrated wave exposure on a power scale (J m^-3^). These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and conditions known to affect coral performance and *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and mediate the supply of food resources that play important roles in coral nutrition [@Houlbreque:2009gh]. SST and chlorophyll were determined using data from the MODIS sensor on the Aqua satellite. SST was evaluated from nighttime records, and chlorophyll from ocean colour, both of which were obtained at 1 km spatial resolution for each month, averaged over the years 2003-2010, from the IMaRS website (http://imars.usf.edu/modis); if data were not available at the chosen coordinates (above), values from the nearest available pixel were used. Boxplots were generated from the 12 monthly climatological values obtained for each response (Fig. 1).

@@

  A total of 84 coral samples representing ten different host species were collected across two sites on the north shore and five sites on the south shore of St. John (Fig. 1). The sequence data generated from these samples comprised 1,490,813 sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.

@@

```{r Fig1, fig.height=3, fig.width=6.85, fig.cap="Figure 1. Sampling of *Symbiodinium* communities from ten different host species (embedded histogram) on the north and south shores of St. John, US Virgin Islands, and the broad-scale physical conditions associated with these two environments. Mapped points along each shore indicate locations from which biological samples were collected, while the 'N' and 'S' points indicate locations from which environmental data representative of the north and south shores were obtained. Boxplots show the sea-surface temperatures (top), chlorophyll A concentrations (middle), and wave exposure (bottom) recorded at these two locations."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

@@

# Plot histogram of sample collections
par(new=T)
samdat <- data.frame(sample_data(phy97))
nsam <- with(samdat, aggregate(Species, by=list(Species=Species, Shore=Shore), FUN=length))
nsam <- dcast(nsam, Species~Shore)
rownames(nsam) <- nsam$Species
rownames(nsam) <- str_replace_all(rownames(nsam), 
                c("alcicornis"="M.alc.", "annularis"="O.ann.", "astreoides"="P.ast.","cavernosa"="M.cav.",
                  "cylindrus"="D.cyl.", "fragum"="F.fra.", "furcata"="P.fur.", "radians"="S.rad.",
                  "siderea"="S.sid.", "strigosa"="P.str."))
nsam <- nsam[,2:3]
par(mar=c(7,5,5,6), mgp=c(0.5,0.2,0))
barplot(t(as.matrix(nsam)), las=2, ylab="# samples", xpd=NA, cex.axis=0.5, cex.names=0.5, cex.lab=0.5, tcl=0.1)
legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1, cex=0.5, legend=c("N","S"), bty="n", inset=c(0.12,-0.1), xpd=T, y.intersp=0.8, x.intersp=0.75)

@@

   The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach used (Table 1 and Fig. S1). More OTUs were resolved as the clustering resolution increased (i.e., 97% OTUs across samples < 97% OTUs within samples < 100% OTUs), with fewer reads per OTU (Table 1). The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n` reads per OTU.

@@

```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A.) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B.) 97% across-sample clustering and C.) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A.) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the taxonomic assignment of the OTU (see Methods)."}
set.seed(42432) 
par(mfrow=c(1,1))
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

@@

  Within the set of coral samples obtained in this study, clade B had the highest relative abundance (`r with(cladeAbund, Prop[Clade=="B"])`%), followed by clade C (`r with(cladeAbund, Prop[Clade=="C"])`%), A (`r with(cladeAbund, Prop[Clade=="A"])`%), D (`r with(cladeAbund, Prop[Clade=="D"])`%), and G (`r with(cladeAbund, Prop[Clade=="G"])`%. The number of OTUs within each clade followed a similar pattern, with the highest number in clade B (`r with(cladeAbund, Notus[Clade=="B"])`), followed by clade C (`r with(cladeAbund, Notus[Clade=="C"])`), A (`r with(cladeAbund, Notus[Clade=="A"])`), D (`r with(cladeAbund, Notus[Clade=="D"])`), and G (`r with(cladeAbund, Notus[Clade=="G"])`). The distribution of these clades within each sample is shown in Fig. 3.

@@

  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any host species (Table 2). However, some qualitative differences were apparent: *Porites furcata* colonies were dominated by either clade A or clade C *Symbiodinium* on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.056). *Siderastrea siderea* was dominated by either clade C or clade D on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.071).

```{r Table2, cache=T } 
set.seed(43789)  
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
shorestats97bs$Species <- str_replace_all(
  shorestats97bs$Species,
  c("alcicornis"="*Millepora alcicornis*",
    "annularis"="*Orbicella annularis*",
    "astreoides"="*Porites astreoides*",
    "cavernosa"="*Montastraea cavernosa*",
    "cylindrus"="*Dendrogyra cylindrus*",
    "fragum"="*Favia fragum*",
    "furcata"="*Porites furcata*",
    "radians"="*Siderastrea radians*",
    "siderea"="*Siderastrea siderea*",
    "strigosa"="*Pseudodiplora strigosa*")
)
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 2. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")

@@

#### Network analyses of *Symbiodinium* metacommunity
```{r networks,  cache=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

@@

  Patterns of association between hosts and *Symbiodinium* were analyzed by constructing three networks focusing on "abundant" (>1% relative abundance; Fig. 4), "dominant" (>50% relative abundance; Fig. 5A), and "background" (<1% relative abundance; Fig. 5B) symbionts in each host species, based on 97% within-sample clustering. (Networks for individuals within each species are presented in Supplementary Materials). In the network for abundant symbionts (Fig. 4), `r sum(V(abunet)$type==2)` OTUs were observed. *P. strigosa* associated with the greatest number of OTUs (n=`r abu.Nedges["strigosa"]`), followed by *F. fragum* (n=`r abu.Nedges["fragum"]`), *M. alcicornis* (n=`r abu.Nedges["alcicornis"]`), and *D. cylindrus* (n=`r abu.Nedges["cylindrus"]`); these primarily comprised OTUs closely related to B1 and B19. In contrast, only a single OTU occurred at >1% relative abundance in both *P. astreoides* (A4) and *M. cavernosa* (C3). All coral species hosted at least one "abundant" *Symbiodinium* OTU that was also abundant in at least one other coral species, except for *S. radians*. This species was dominated by C46, but also contained *Symbiodinium* closely related to B1 and B19, as well as B5.

@@

  There were `r sum(V(bkgnet3)$type==2)` symbionts that occupied a 'background' niche, defined as <1% relative abundance in three or more coral species (Fig. 5B). A member of *Symbiodinium* clade D (which has been formally described as *S. trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species examined (`r nspp["D1"]`), followed by A4 and C3 (both were in `r nspp["C3"]` taxa). C31, C46, and B1 were each detected at background levels in 4 host species.

```{r Fig4, fig.width=5, fig.height=5, fig.align="center", fig.retina=1, out.width=5, fig.cap="<br>Figure 4. Network analysis of abundant *Symbiodinium* taxa in each coral species. *Symbiodinium* OTUs (circular nodes) are connected to each coral species (square nodes) in which they ever occurred at >1% relative abundance within an individual; thickness of edges (i.e., links between coral and symbiont nodes) is relative to the proportion of individuals within the coral species in which the *Symbiodinium* OTU was present at >1%. Symbiont nodes are colored by clade identity, and sized proportionally to the number of coral species they dominated."}

@@

  In the ten host species sampled in St. John, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324]. Less frequent associations between hosts and members of clades A and D were observed, although clade A dominated *P. astreoides* and some *P. furcata* colonies, and clade D dominated one *S. siderea* colony. Finally, clade G was observed at low relative abundances (<`r 100*round(clademax[5,2], 2)`%) in some samples, which likely would not have been detected without high-throughput sequencing.

>>>>

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance and stress-tolerance of their coral hosts, making the analysis of *Symbiodinium* communities in corals (and metacommunities on reefs) advantageous for many aspects of coral reef research. High-throughput sequencing of ITS2 nrDNA offers unprecedented scale in describing these communities, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity. Here, we demonstrate that generating operational taxonomic units by clustering ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific. We utilize this 'within-sample clustering' to analyze *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St. John, US Virgin Islands. While *Symbiodinium* communities did not differ between shores, metacommunity network analysis of host-symbiont associations revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are more 'flexible' or 'specific' in their associations with *Symbiodinium*. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

@@

  *Symbiodinium* identity is primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; @Santos:2003kt; @Pochon:2012we]. Together, these markers have been used to identify nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which have been further divided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, ecological surveys of *Symbiodinium* diversity must still rely on commonly-used marker genes, such as ITS2. Consequently, high-throughput sequencing of ITS2 is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs]. However, such analyses may create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within the genus [@Finney:2010p8324; @Thornhill:2013dx]. For example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially other species [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly-repeating ribosomal DNA array creates multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (e.g. [@Dover:1986ju]). In fact, concerted evolution may mask species divergence within *Symbiodinium* by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker for *Symbiodinium*. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level. Moreover, due to the large quantity of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2, it remains an essential and powerful marker for *Symbiodinium*. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys of *Symbiodinium*.

  Here, we describe a *Symbiodinium* metacommunity associated with scleractinians and a *Millepora* hydrocoral in St. John, U.S. Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that can advance understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' (i.e. clustering sequences at 97% similarity within each sample independently) to generate biologically relevant operational taxonomic units (OTUs) from ITS2 metabarcoding data. Specifically, this approach addresses the fact that dominant ITS2 sequences from different *Symbiodinium* species may be more similar to one another than intragenomic variants within one *Symbiodinium* [@Thornhill:2013dx; @Arif:2014gk; @Parkinson:2015gt]. Therefore, clustering all sequences together underestimates diversity by collapsing different species into the same OTU. Conversely, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to intragenomic variation. Within-sample clustering may better approximate true diversity by exploiting key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts. These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to be intragenomic variants, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa. Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using within-sample clustering, we analyzed *Symbiodinium* communities in ten host species across the north and south shores of St. John. We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying the variability in symbiont communities (i.e., beta diversity, or 'symbiosis flexibility') among individuals within a host species. The large quantity of data, and the kinds of analytical approaches facilitated by ITS2 metabarcoding, have the potential to revolutionize understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts. More generally, the present study demonstrates how ecological knowledge can inform bioinformatic analyses using marker genes for which sequence diversity does not map directly to species.

@@

  Eighty-four tissue samples were collected from scleractinian and milleporine corals at seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and colonies to be sampled were selected haphazardly in upward-facing locations exposed to ambient irradiance. Tissue biopsies (~5-mm diameter and ~5-mm depth) were removed from coral colonies with metal borers, and the holes were plugged with non-toxic modeling clay.

  Broad-scale differences in the hydrographic regimes at points representative of the north and south shores of St. John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and integrated wave exposure on a power scale (J m^-3^). These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and conditions known to affect coral performance, *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and the supply of food resources that play important roles in coral nutrition [@Houlbreque:2009gh]. SST and chlorophyll were determined using data from the MODIS sensor on the Aqua satellite. SST was evaluated from nighttime records, and chlorophyll from ocean colour, both of which were obtained at 1 km spatial resolution for each month, averaged over the years 2003-2010, from the IMaRS website (http://imars.usf.edu/modis); if data were not available at the chosen coordinates (above), values from the nearest available pixel were used. Boxplots were generated from the 12 monthly climatological values obtained for each response (Fig. 1).

@@

  A total of 84 coral samples representing ten host species were collected at two sites on the north shore and five sites on the south shore of St. John (Fig. 1). These samples generated 1,490,813 sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.

@@

```{r Fig1, fig.height=3, fig.width=6.85, fig.cap="Figure 1. Sampling of *Symbiodinium* communities from ten host species (embedded histogram) on the north and south shores of St. John, US Virgin Islands, and the broad-scale physical conditions associated with these two environments. Mapped points along each shore indicate locations from which biological samples were collected, while 'N' and 'S' indicate locations from which environmental data representative of the north and south shores were obtained. Boxplots show the sea-surface temperatures (top), chlorophyll *a* concentrations (middle), and wave exposure (bottom) recorded at these two locations."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

@@

# Plot histogram of sample collections
par(new=T)
samdat <- data.frame(sample_data(phy97))
nsam <- with(samdat, aggregate(Species, by=list(Species=Species, Shore=Shore), FUN=length))
nsam <- dcast(nsam, Species~Shore)
rownames(nsam) <- nsam$Species
rownames(nsam) <- str_replace_all(rownames(nsam), 
                c("alcicornis"="M.alcicornis", "annularis"="O.annularis", "astreoides"="P.astreoides","cavernosa"="M.cavernosa",
                  "cylindrus"="D.cylindrus", "fragum"="F.fragum", "furcata"="P.furcata", "radians"="S.radians",
                  "siderea"="S.siderea", "strigosa"="P.strigosa"))
nsam <- nsam[,2:3]
par(mar=c(7,5,5,6), mgp=c(0.5,0.2,0))
b <- barplot(t(as.matrix(nsam)), las=2, ylab="# samples", xpd=NA, names.arg=rep("",10), cex.axis=0.5, cex.names=0.5, cex.lab=0.5, tcl=0.1)
text(b+1.2, -0.5, labels=rownames(nsam), srt=45, pos=2, cex=0.5, adj=1)
legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1, cex=0.5, legend=c("N","S"), bty="n", inset=c(0.12,-0.1), xpd=T, y.intersp=0.8, x.intersp=0.75)

@@

   The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach (Table 1 and Fig. S1). More OTUs were resolved as the clustering resolution increased (i.e., 97% OTUs across samples < 97% OTUs within samples < 100% OTUs), with fewer reads per OTU (Table 1). The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n` reads per OTU.

@@

```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B) 97% across-sample clustering and C) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the taxonomic assignment of the OTU (see Methods)."}
set.seed(42432) 
par(mfrow=c(1,1))
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

@@

  Within the set of coral samples, clade B had the highest relative abundance (`r with(cladeAbund, Prop[Clade=="B"])`%), followed by clade C (`r with(cladeAbund, Prop[Clade=="C"])`%), A (`r with(cladeAbund, Prop[Clade=="A"])`%), D (`r with(cladeAbund, Prop[Clade=="D"])`%), and G (`r with(cladeAbund, Prop[Clade=="G"])`%). The number of OTUs within each clade followed a similar pattern, with the highest number in clade B (`r with(cladeAbund, Notus[Clade=="B"])`), followed by clade C (`r with(cladeAbund, Notus[Clade=="C"])`), A (`r with(cladeAbund, Notus[Clade=="A"])`), D (`r with(cladeAbund, Notus[Clade=="D"])`), and G (`r with(cladeAbund, Notus[Clade=="G"])`). The distribution of these clades within each sample is shown in Fig. 3.

@@

  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any host species (Table 2). However, qualitative differences were apparent: *Porites furcata* was dominated by either clade A or clade C *Symbiodinium* on the south shore, but only by clade C on the north shore (difference between shores PERMANOVA: p=0.056). *Siderastrea siderea* was dominated by either clade C or clade D on the south shore, but only by clade C on the north shore (difference between shores PERMANOVA: p=0.071).

```{r Table2, cache=T } 
set.seed(43789)  
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
shorestats97bs$Species <- str_replace_all(
  shorestats97bs$Species,
  c("alcicornis"="*Millepora alcicornis*",
    "annularis"="*Orbicella annularis*",
    "astreoides"="*Porites astreoides*",
    "cavernosa"="*Montastraea cavernosa*",
    "cylindrus"="*Dendrogyra cylindrus*",
    "fragum"="*Favia fragum*",
    "furcata"="*Porites furcata*",
    "radians"="*Siderastrea radians*",
    "siderea"="*Siderastrea siderea*",
    "strigosa"="*Pseudodiplora strigosa*")
)
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 2. Mean overall, within-shore ('within'), and between-shore ('between') Bray-Curtis dissimilarities of the *Symbiodinium* communities in each host species, and PERMANOVA tests (partial R^2^ and p-values) for a difference between shores. Between-shore tests were not possible for *O. annularis* and *P. astreoides* since they were only sampled from one shore.")

@@

#### Network analysis of *Symbiodinium* metacommunity
```{r networks,  cache=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

@@

  Patterns of association between hosts and *Symbiodinium* were analyzed using networks focusing on "abundant" (>1% relative abundance; Fig. 4), "dominant" (>50% relative abundance; Fig. 5A), and "background" (<1% relative abundance; Fig. 5B) symbionts in each host species, based on 97% within-sample clustering. (Networks for individuals within each species are presented in Supplementary Materials). In the network for abundant symbionts (Fig. 4), `r sum(V(abunet)$type==2)` OTUs were observed. *P. strigosa* associated with the greatest number of OTUs (n=`r abu.Nedges["strigosa"]`), followed by *F. fragum* (n=`r abu.Nedges["fragum"]`), *M. alcicornis* (n=`r abu.Nedges["alcicornis"]`), and *D. cylindrus* (n=`r abu.Nedges["cylindrus"]`); these primarily comprised OTUs closely related to B1 and B19. In contrast, only a single OTU occurred at >1% relative abundance in both *P. astreoides* (A4) and *M. cavernosa* (C3). All coral species hosted at least one "abundant" *Symbiodinium* OTU that was also abundant in at least one other coral species, except for *S. radians*. This species was dominated by C46, but also contained *Symbiodinium* closely related to B1 and B19, as well as B5.

@@

  There were `r sum(V(bkgnet3)$type==2)` symbionts that occupied a 'background' niche, defined as <1% relative abundance in three or more coral species (Fig. 5B). A member of *Symbiodinium* clade D (which has been described as *S. trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species (`r nspp["D1"]`), followed by A4 and C3 (both in `r nspp["C3"]` taxa). C31, C46, and B1 were each detected at background levels in 4 host species.

```{r Fig4, fig.width=5, fig.height=5, fig.align="center", fig.retina=1, out.width=5, fig.cap="<br>Figure 4. Network analysis of abundant *Symbiodinium* taxa in each coral species. *Symbiodinium* OTUs (circular nodes) are connected to each coral species (square nodes) in which they ever occurred at >1% relative abundance within an individual; thickness of edges (i.e., links between coral and symbiont nodes) is relative to the proportion of individuals within the coral species in which the *Symbiodinium* OTU was present at >1%. Symbiont nodes are colored by clade identity, and sized proportionally to the number of coral species in which they were found."}

@@

  In the ten host species sampled, the most prevalent *Symbiodinium* were members of clades B and C, which is consistent with previous analyses of *Symbiodinium* diversity on shallow reefs in this region [@LaJeunesse:2004p122; @Correa:2009p4378; @Finney:2010p8324]. Less frequent associations between hosts and members of clades A and D were observed, although clade A dominated *P. astreoides* and some *P. furcata*, and clade D dominated one *S. siderea*. Finally, clade G was observed at low relative abundances (<`r 100*round(clademax[5,2], 2)`%) in some coral colonies, which likely would not have been detected without high-throughput sequencing.
